---
title: "Real-world Progression-free Survival (rwPFS)"
author: "Michael Bretscher"
date: "`r Sys.Date()`"
bibliography: references.bib 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-world Progression-free Survival (rwPFS)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# knitr::opts_chunk$set(
#   collapse = TRUE,
#   comment = "#>",
#   fig.height = 5,fig.width = 7
# )


library(knitr)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,  # set to TRUE for running vignette (but must remove dependence on FK)
  echo = TRUE,         # echo code?
  message = TRUE,     # Show messages
  warning = TRUE,     # Show warnings
  fig.width = 8,       # Default plot width
  fig.height = 6,      # .... height
  dpi = 200,           # Plot resolution
  fig.align = "center"
)


#Run this code to build vignettes & make them accessible through R help - also when installing from git
# tools::buildVignettes(dir = ".", tangle=TRUE)
# dir.create("inst/doc")
# file.copy(dir("vignettes", full.names=TRUE), "inst/doc", overwrite=TRUE)
# ..and push the changes to git..
#From: https://community.rstudio.com/t/browsevignettes-mypackage-saying-no-vignettes-found/68656/7

```


*The background section gives some general context on using real-world Progression-free Survival as endpoint in oncology - if you just want to know about using the RwPFS package feel free to skip ahead to [Getting started with RwPFS](#getting-started-with-rwpfs)* 

<br/>
*For a detailed discussion of the rwPFS endpoint see @rwPFS_OAK*
<br/>
<br/>


# Background on rwPFS


Progression-free Survival (PFS) is a commonly used time-to event endpoint in cancer trials. It is defined as the first event of either death or progression.

An important difference between the PFS and Overall Survival (OS, time to death) endpoints is that PFS requires close follow-up of patients, whereas OS does not. 
This is because death is a one-time event, and whenever a patient is contacted alive, we know that death had not occurred previously. This is not the case when looking at progression, hence patients must be closely monitored.

<br/>

**The mechanisms of data-collection differ between real-world data & clinical trials [@rwPFS_OAK]:**

- In clinical trials, tumor assessments are scheduled at regular intervals, and progression is assessed using standardized criteria. The reason for collecting PFS data in clinical trials is to measure treatment efficacy. Transparency and consistency of the data collection process are emphasized.

- In routine health care, there is commonly no protocol and no regularly scheduled tumor assessments. Progression may or may not be assessed following standard criteria - this is generally not known. Physicians may measure progression mainly in order to make treatment decisions, not to statistically compare the effectiveness of different treatments. In addition, transparency and consistency of data collection are likely to be less important objectives than cost considerations or minimization of discomfort for the patient (e.g. by avoiding unnecessary scans).


<br/>

Because of these differences, it appears sensible to speak of "real-world PFS" (**rwPFS**) and to treat it as a separate endpoint, distinct from trial PFS. However, due to the highly similar definitions of rwPFS and PFS, it is possible that the two endpoints may be *equivalent*, meaning they essentially lead to the same result and could be used to compare effectivness of treatments both within a particular real-world data source as well as between trial and real-world data.


<br/>

> Whether or not **PFS** and **rwPFS** endpoints are **equivalent** and can be used for comparing drug efficacy across two data sets (real-world and 
> clinical trial) may depend on type of cancer, drug category, real-world database, and other factors. While attempts should be made to 
> understand and ensure quality of data collection in the real-world data, this will likely never be perfect. Consequently, empirical confirmation
> of the equivalence of rwPFS and PFS will be needed. This can be done by replicating trial arms and comparing rwPFS and trial PFS outcomes 
> [@rwPFS_OAK; @EC_impowers], across indications and drug categories.



## Algorithm for calculating rwPFS using Flatiron data 


Figure 1 in @rwPFS_OAK illustrates the important elements required for calculating a rwPFS endpoint:

- the start date (baseline) 

- the date up to which progression was abstracted from clinical notes

- the dates of any progression events of interest

- the date of death, if present

- the date of a patient's last structured activity in the database

- the presence of potential visit gaps (not shown), indicating that the patient may not have been followed for progression due to prolonged absence from the clinic in the Flatiron network (e.g. >90d). 

- a time window after the end of progression follow-up where death events are to be captured. This is required to avoid informative censoring because it is not uncommon that data on progression ceases to be collected shortly before death of a patient. If the time window is too small, there will almost exclusively be progression events. If the window is too large, more deaths are captured, but at the same time the risk of missing any progression events increases. Therefore, a pragmatic choice in the middle will have to be made. More on this point later.


<br/>
The RwPFS package uses the algorithm below to calucate rwPFS endpoints.
<br/>  
  

>**A general algorithm for calculation of rwPFS consists of the following steps:**
>
>1. **Determine the period of continuous progression follow-up.**
>    
>     It begins at baseline and ends at one of the following, whichever occurs first:
>    
>       - the start of a long gap between visits (e.g. >90d)
>   
>       - the date of last progression abstraction
>   
>       - the date of last structured activity in the database
>  
>
>
>2. **Determine the type of rwPFS event**, in this order:
>
>     - "Progression", if an event occurs during the follow-up period
> 
>     - "Death", if it occurs within the specified time window after end of follow-up
> 
>     - "Censoring" otherwise
>
>  
>  
>
>3. **Create rwPFS variables **
>
>     - record a rwPFS event in case of death or progression event types
>
>     - otherwise, censor the patient at the end of progression follow-up
>
>     - calculate the time from baseline to event or censoring
  
  


## There are several possible rwPFS definitions
  
  

**Multiple possible definitions of rwPFS can be specified based on the elements mentioned above:** 
  
  
  - **different subsets of progression events could be used**, e.g. only radiographically confirmed progression events, or only progression events that occur more than 14 days after baseline
  
  - **different possibilities exist to handle changes in the line of therapy (LoT):** one could either
  
    - follow the patient for progression into a subsequent LoT if no progression has occurred, or..
    
    - censor the patient at the start of a subsequent LoT, or..
    
    - impute a progression event at the start of a subsequent LoT (following the clinical logic that LoT is anchored to progression)
    
  - **different choices are possible for the length of the time window to capture death events**
  
    - the aim is to capture most of - and only - the death events correlated in time with the end of follow-up.
    
    - this is reminiscent of informative censoring - except in this case events occurring after the *end of follow-up* can be accessed and potentially included.
    
    - reasonable values for the length of the time window can be argued based on comparison with data collection in trials, or - in a data-driven way by inspecting the drop in the number of death events as the window is being extended (see best practices below). For a detailed discussion see @rwPFS_OAK 
    
    - ultimately, any choice will remain arbitrary to some extent.
    
  - it seems reasonable to ensure gapless progression follow-up, but **the choice of a maximally tolerable gap length between visits remains arbitrary**. A commonly used value is 90 days, however, without a very strong rationale.
    
    
    
## Best practices to support the emergence of consensus definitions   
    
As outlined above, there is a range of possible definitions of rwPFS. To determine which of these are most closely mimicking clinical trial PFS requires discussions with subject matter experts, and it is possible that different conclusions may be reached depending on cancer type, indication, drug category, analysis objective, or other factors. From our experience in aNSCLC [@rwPFS_OAK; @EC_impowers] several definitions - not only one - remain viable options even after discussion with experts, and eventually a pragmatic choice must be made. Despite a desire for standardization, we reached the conclusion that defining a single standard is not helpful given the current state of knowledge, and that instead we should promote best practices that support the emergence of - preferably rather few -  standard definitions via publications in the scientific literature.

<br/>

To promote the emergence of standard rwPFS definitions and comparability across analyses we propose the following best practices:
<br/>
<br/>

>**Proposed best practices when using rwPFS**
>
> 1. Check if there is already a commonly used rwPFS definition in analyses similar to yours.
>
> 2. Discuss sensible rwPFS definitions with subject matter experts.
>
> 3. If there is already a commonly used rwPFS definition that passes the expert test, try to use it.
>
> 4. If you have reasons to use a different definition, do so and try to explain your choice
>
> 5. **Be transparent about the exact definition you're using**
>
> 6. **Explore a range of sensible alternative definitions and report key descriptors to show how rwPFS definition affects the conclusions of the analysis**
  
  
<br/>
<br/>

TODO: continue here

# Getting started with RwPFS
  
  
To start, we will add a rwPFS variable to our FlatironKitchen dataset. For simplicity, we will use the rwPFS definition used in @rwPFS_OAK and @EC_impowers:
  
  
- **Use all progression events except where “pseudoprogression mentioned”** (not only events based on radiographic evidence, and we don’t discard progression events occurring within 2 weeks after baseline)

- **Follow patients for progression into subsequent lines of treatment** (do not censor them at start of new line, do not impute progression at line change)

- **Include death events occurring <30d after end of progression follow-up**

- **Censor individuals if there is a visit gap > 90 days during follow-up**

- **Use the first progression event when there are several**
  
  

## Setup

1. Load the required libraries

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(FlatironKitchen)
```

Here is the version of `FlatironKitchen` used in this vignette.

```{r package_version}
packageVersion("FlatironKitchen")
```



2. Create a FlatironKitchen object, using the advanced Non Small-cell Lung Cancer datamart 

```{r, results = FALSE}
fk <- fi_start(
  datamart = "AdvancedNSCLC",
  title = "rwPFS in aNSCLC"
) 
```

3. Add information on the 1st line of therapy (note we're also getting info on the 2nd line for later..)

```{r, results = FALSE, message = FALSE}
fk <- fk %>%  
  fi_add_lineoftherapy_flatiron(
    lines = c(1,2),
    index_date = "advanceddiagnosisdate",
    left = 0,
    right = 90,
    calc_duration = FALSE
  ) 
```

4. We will look at the rwPFS in patients that receive Carboplatin & Paclitaxel in first line

```{r, results = FALSE}
fk <- fk %>%  
  fi_cohort_include(
    lot1linename == "Carboplatin,Paclitaxel", 
    description = "Carboplatin & Paclitaxel in 1st line", 
    keep_na = FALSE
  ) 
```


Let's a have a quick look at the cohort we've just selected:

```{r}
print(fk)
```





These are the variables we currently have in our data:

```{r}
fk$data %>% names
```

Note that several variables required for the calculation of rwPFS are already present. 

These are "lot1startdate" (our baseline), lastactivitydate and dateofdeath.

What's still missing for the calculation of rwPFS (besides progression), is information on any long gaps between visits, so let's do this


5. Add information on visit gaps larger than 90 days from start of 1st line onwards


```{r, message = FALSE, results = FALSE}
fk <- fk %>%  
  fi_calc_visitgap(
    index_date = "lot1startdate",
    gapdays = 90,
    what = "After",
    force_database = FALSE
  )
```

This is required to ensure gapless progression follow-up. The function added a few columns related to the visit gap, of which we only need the date of the "lastvisitbeforegap".

## Add information on progression

6. add the date of the first progression event and end date of progression abstraction ("lastclinicnotedate")

Selecting the subset of progression events to use is the first step of specifying a rwPFS definition. Here we stick to the definition outlined above, using all progression events, radiographic or not, and we don't discard any events occuring right after baseline. We do exclude events that are flagged "ispseudoprogressionmentioned", though. Incidentally, there won't be any pseudoprogression events in chemotherapies such as the Carboplatin & Paclitaxel combination (you could check the number of events to verify). Here we're simply sticking to the definition proposed by @rwPFS_OAK for use across both CIT and chemotherapies in aNSCLC, based on the OAK trial (see @rwPFS_OAK for details).

```{r}
fk <- fk %>% 
  fi_add_progression(
      start_date = "lot1startdate",          #we're measuring rwPFS from the start of treatment
      require_radiographic = FALSE,          #we don't require progression to be radiographically confirmed (in NSCLC, most are anyway)
      exclude_pseudoprogression = TRUE,      #we exclude pseudoprogression events (there won't be any in our chemo treatment)
      discard_n_days = 0,                    #we do not discard any progression events right after baseline              
      label = "_nopseudo",                   #this (optional) label will help us keep different rwPFS endpoints apart later
      prog_filter_expression = NULL          #this is an advanced feature allowing more flexible event selection, explained later. Defaults to NULL.
  )
```

Two new columns were added: *progressiondate_nopseudo* and *lastclinicnotedate_nopseudo*

Note that FlatironKitchen warned us that 287 out of our 7,823 patients had no progression abstracted beyond the baseline date. This can happen if patients had no progression information abstracted at all, or their lastclinicnotedate is before our startdate. At this point we could exclude those patients since it won't be possible to calculate rwPFS for them. However, let's leave them for now, in order to demonstrate how the downstream functions deal with such cases.



## Habemus real-world PFS!

Not yet - but soon. Now that all required columns are present in our dataset, we can..

7. Compute the rwPFS endpoint

```{r}

fk <- fk %>% 
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          #we're measuring rwPFS from the start of treatment
    visit_gap_start_date = "lastvisitbeforegap",           #the start date of any gap in visits longer than 90 days
    last_clinic_note_date = "lastclinicnotedate_nopseudo", #the lastclinicnotedate we just added..
    progression_date = "progressiondate_nopseudo",         #..and the progression date with the same label
    last_activity_date = "lastactivitydate",               #the date of last activity in the database
    death_date = "dateofdeath",
    death_window_days = 30,                                #include death events up to 30d after end of progression follow-up
    max_follow_up_days = Inf,                              #censor patients after a maximum time (e.g. to harmonize with the trial). Defaults to Inf.
    label = "_no_pseudoprogression"                        #let this be our primary rwPFS definition, the reference to compare others against
  )


```
  
  
Note the warning about some patients having been set to missing. Let's have a closer look at those.
  
  
```{r}
fk$data %>% 
  filter(rwpfs_no_pseudoprogression_event_type == "Missing") %>% 
  count(
    `missing lastclinicnotedate` = is.na(lastclinicnotedate_nopseudo), 
    `lastclinicnotedate on start date?` = lot1startdate == lastclinicnotedate_nopseudo,
    `missing last activity date?` = is.na(lastactivitydate),
    `death before start date` = dateofdeath <= lot1startdate
  ) %>%
  knitr::kable() 
```
  
  
It looks like a total of 278 patients had a missing lastclinicnote date (this may have been so from the start, or fi_add_progression may have set this to missing in case it was before the start date). The fi_add_progression function had already alerted us about this. An additional 17 patients seem to have their death dates before start date, which is likely due to the monthly granularity of death dates in Flatiron (for de-identification purposes). Let us remove patients with missing rwPFS.
  
  
```{r}
fk <- fk %>%
  fi_cohort_exclude(rwPFS_no_pseudoprogression_event_type == "Missing",
                    description = "Exclude missing rwPFS",
                    keep_na = FALSE
  )

```


Finally, let's plot the Kaplan-Meier curve for rwPFS, stratified by histology.

```{r, out.width = "100%"}
fk %>%
  fi_stat_survfit(
    followup_duration = "rwpfs_no_pseudoprogression_months",
    status = "rwpfs_no_pseudoprogression_event",
    strata = "histology"
  ) %>%
  fi_plot_km(
    xlab = "Months"
  )
```
  
  
This doesn't look too bad - except that the different histologies have very different follow-up times, it seems.
Let's re-generate those rwPFS variables, but this time right-censor patients with respect to rwPFS at around 40 months - let's say 1200 days.
  
  
```{r}
fk <- fk %>% 
  fi_select(-contains("rwpfs")) %>%                        #FK doesn't let us overwrite columns, lots remove the old ones first.. 
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo", 
    progression_date = "progressiondate_nopseudo",        
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 30,                                
    max_follow_up_days = 1200,                             
    label = "_no_pseudoprogression"                                  
  ) 
```
  
  
..and plot it again
  
  
```{r, out.width = "100%"}
fk %>%
  fi_stat_survfit(
    followup_duration = "rwpfs_no_pseudoprogression_months",
    status = "rwpfs_no_pseudoprogression_event",
    strata = "histology"
  ) %>%
  fi_plot_km(
    xlab = "Months"
  )
```
  
  
This looks much cleaner. Congratulations, you've mastered the basics of using rwPFS!

<br/>
<br/>


# Advanced: comparison of multiple rwPFS definitions

As mentioned before, there is more than one way of defining rwPFS. Often, the differences between definitions will be small - but nevertheless, we need to show this, we can't just assume they are, and we must be able to answer questions in that direction.. However, conducting a full sensitivity analysis for each and every rwPFS definition seems overkill.. so what do we do?

<br/>

Ideally, we'd have a concise summary table, where readers could verify at one glance what the impact of varying the endpoint definition would be.
This is what we'll try to do..

<br/>

First, let's calculate rwPFS across a range of "reasonable" definitions, starting with the progression part & varying one aspect at a time, always comparing against our reference progression variable defined previously. 

<br/>

```{r}
fk <- fk %>%
  
  #do not remove pseudoprogression events
  fi_add_progression(
    start_date = "lot1startdate",          
    require_radiographic = FALSE,          #we don't require progression to be radiographically confirmed (in NSCLC, most are anyway)
    exclude_pseudoprogression = FALSE,     #we don't exclude pseudoprogression events (there won't be any in our chemo treatment)
    discard_n_days = 0,                    #we do not discard any progression events right after baseline              
    label = "_allevents",                  
    prog_filter_expression = NULL          
  ) %>%
  
  #omit progression events occurring within 14d of baseline (Flatiron guidance)
  fi_add_progression(
    start_date = "lot1startdate",          
    require_radiographic = FALSE,          #we don't require progression to be radiographically confirmed (in NSCLC, most are anyway)
    exclude_pseudoprogression = TRUE,      #we exclude pseudoprogression events (there won't be any in our chemo treatment)
    discard_n_days = 14,                   #we discard any progression events within 14d after baseline              
    label = "_omit_le_14d",                
    prog_filter_expression = NULL         
  ) %>%
  
  #only use radiographically confirmed progression events
  fi_add_progression(
    start_date = "lot1startdate",          
    require_radiographic = TRUE,           #we require progression to be radiographically confirmed (in NSCLC, most are anyway)
    exclude_pseudoprogression = TRUE,      #we exclude pseudoprogression events (there won't be any in our chemo treatment)
    discard_n_days = 0,                    #we don't discard progression events shortly after baseline              
    label = "_only_radiographic",          
    prog_filter_expression = NULL   
  ) 
  


```

  
<br/>

We also want to look at rwPFS definitions where

- a progression event is imputed at the start of the next LoT 

- patients are censored at start of the next Lot

- we don't consider visit gaps
  
<br/>

We already have the start date of the 2nd LoT from before. What we need is the date of those imputed progression events, and a columns that replaces the lastclinicnotedate with the start date of LoT 2 if it's before lastclinicnotedate.. 
  
<br/>
  
```{r}
fk <- fk %>%
  fi_mutate(
    #impute progression at start of 2nd line
    imputed_progression_date = pmin(progressiondate_nopseudo, lot2startdate, na.rm = T),
    #censor at the earliest of lastclinicnotedate or start of 2nd line
    imputed_censoring_date = pmin(lastclinicnotedate_nopseudo, lot2startdate, na.rm = T),
    #adding an all_na column because fi_calc_rwPFS expects the start date of any long visit gaps - we'll tell it there are none..
    all_na = NA_character_ %>% as.Date()
  )
  


```

<br/>

Now, we can calculate the different rwPFS endpoints corresponding to the above handling of progression events and LoT:

<br/>

```{r}
fk <- fk %>% 
  
   #all events
   fi_calc_rwPFS(
    start_date = "lot1startdate",                           
    visit_gap_start_date = "lastvisitbeforegap",            
    last_clinic_note_date = "lastclinicnotedate_allevents",
    progression_date = "progressiondate_allevents",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 30,                                 
    max_follow_up_days = 1200,                              
    label = "_all_events"                                   
  )  %>%
  
  #omit events within 14d of baseline
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",          
    last_clinic_note_date = "lastclinicnotedate_omit_le_14d", 
    progression_date = "progressiondate_omit_le_14d",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 30,                                
    max_follow_up_days = 1200,                             
    label = "_omit_le_14d"                                
  ) %>%
  
  #only use radiographically confirmed progression events
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_only_radiographic", 
    progression_date = "progressiondate_only_radiographic", 
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 30,                                
    max_follow_up_days = 1200,                             
    label = "_only_radiographic"                           
  )   %>%
  
  #impute progression at start of 2nd line
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo", 
    progression_date = "imputed_progression_date",        
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 30,                                
    max_follow_up_days = 1200,                             
    label = "_impute_progression"                          
  ) %>%
  
  #censor at the earliest of lastclinicnotedate or start of 2nd line
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "imputed_censoring_date",      
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 30,                                
    max_follow_up_days = 1200,                             
    label = "_censor_at_lot_change"                        
  ) %>%
  
  #Ignore visit gaps
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "all_na",                       
    last_clinic_note_date = "lastclinicnotedate_nopseudo", 
    progression_date = "progressiondate_nopseudo",        
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 30,                                
    max_follow_up_days = 1200,                             
    label = "_ignore_visit_gaps"                           
  ) 
  
```
<br/>

Now that we have all these different rwPFS endpoints, let's compare them using `fi_compare_rwPFS`: 

<br/>
```{r}

fk %>% 
  fi_compare_rwPFS(
    labels = c(                         #Specify the rwPFS labels to be tabulated (in the desired order)
      "_no_pseudoprogression",
      "_all_events",
      "_omit_le_14d",
      "_only_radiographic",
      "_impute_progression",
      "_censor_at_lot_change",
      "_ignore_visit_gaps"
    ),
    reference = "_no_pseudoprogression",
    incremental_deaths_column = FALSE 
  ) %>%
  kable()

```

<br/>


The obtained table gives an overview across the rwPFS definitions we're considering. It compares the no. of patients censored, with progression, or dying, as well as the percentage of rwPFS events (death and progression combined) that are death. 
  
  
Kaplan-Meier median and hazard ratio relative to the reference definition give an impression of how big the impact on corresponding analysis outcomes would be. Note that to obtain the hazard ratio, a cox regression was fitted to data of twice the same patients but with different endpoints. Since the Cox model assumes the data from two compared groups to be independent, this is a bit of a stretch. The HR should therefore cautiously be interpreted as a crude indication of expected impact on any HR analyses, in order to avoid doing full senstivity analyses with all rwPFS definitions.
  
  
Glancing at HR's across all definitions, we find that in most cases, hazard ratios should be little affected by the exact rwPFS definition, with two exceptions: imputing progression increases the event rate considerably, and not ensuring continuous progression follow-up slightly reduces the event rate, presumably due to progression events going missing. 
  
  
With 'imputing progression' having the strongest impact, arguing for this definition would require strong evidence that there is a substantial amount of missing information in the progression data, and that line-change would be a reliable indicator of progression. We consider this not to be the case and conclude that imputing progression at line change is not to be recommended. 
  
  
Not ensuring continuous follow-up by not censoring patients at the start of any visit gap of 90d or longer decreases the event rate, which suggests that indeed the progression information in patients with long gaps is incomplete. Consequently, we consider it reasonable to censor patients at the start of any long gaps, as is done in the reference definition.
  
  
Apart from 'Ignore visit gaps' and 'Impute progression', all tested definitions appear sensible and differ very little in terms of hazard ratios or median rwPFS.

<br/>

A remaining question is how long the time window for capturing deaths should be, after the end of progression follow-up. We've chosen 30 days above, 
but mostly because that's what @rwPFS_OAK and @EC_impowers did. 
  
<br/>
  
There may not be one single "right" value for the size of this window. However, one way of verifying if a given value is sensible is to vary the length
of this window and check how many additional deaths are captured when extending it. The idea here is that we want to capture most of the deaths that are
correlated with the end of progression follow-up, meaning they happen shortly after. In addition, we want to keep the window as small as possible to
avoid missing any progression events. A pragmatic choice may be made that reconciles these opposing objectives as well as possible.
  
<br/>  
We can calculate rwPFS endpoints with varying window lengths as follows:
  
<br/>  

```{r}
fk <- fk %>%
  
  #removing all previous rwPFS columns
  fi_select(-contains("rwpfs")) %>%     
  
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo",
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 0,                                
    max_follow_up_days = 1200,                             
    label = "_0d_window"                                   
  ) %>%
  
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo",
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 10,                                
    max_follow_up_days = 1200,                             
    label = "_10d_window"                                   
  ) %>%
  
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo",
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 20,                                
    max_follow_up_days = 1200,                             
    label = "_20d_window"                                   
  ) %>%
  
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo",
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 30,                                
    max_follow_up_days = 1200,                             
    label = "_30d_window"                                   
  ) %>%
  
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo",
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 40,                                
    max_follow_up_days = 1200,                             
    label = "_40d_window"                                   
  ) %>%
  
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo",
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 50,                                
    max_follow_up_days = 1200,                             
    label = "_50d_window"                                   
  ) %>%
  
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo",
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 60,                                
    max_follow_up_days = 1200,                             
    label = "_60d_window"                                   
  ) %>%
  
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo",
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 70,                                
    max_follow_up_days = 1200,                             
    label = "_70d_window"                                   
  ) %>%
  
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo",
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 80,                                
    max_follow_up_days = 1200,                             
    label = "_80d_window"                                   
  ) %>%
  
  fi_calc_rwPFS(
    start_date = "lot1startdate",                          
    visit_gap_start_date = "lastvisitbeforegap",           
    last_clinic_note_date = "lastclinicnotedate_nopseudo",
    progression_date = "progressiondate_nopseudo",         
    last_activity_date = "lastactivitydate",               
    death_date = "dateofdeath",
    death_window_days = 90,                                
    max_follow_up_days = 1200,                             
    label = "_90d_window"                                   
  ) 
 
  
```
  
<br/>

and then compare them like before, but this time adding an 'incremental deaths' column to the output, indicating how many additional deaths are captured by extending the window by 10 more days

<br/> 
  
```{r}
fk %>% 
  fi_compare_rwPFS(
    labels = paste0("_", 0:9*10, "d_window"),                         #Specify the rwPFS labels to be tabulated (in the desired order)
    reference = "_30d_window",
    incremental_deaths_column = TRUE 
  ) %>%
  kable()

```
  
<br/>

Similarly as before, we are now able to compare different rwPFS definitions at a glance and examine the impact of extending the time window for capturing death events.
  
<br/>

First, a time-window of 0 days corresponds to a reasonably substantial HR difference of 0.2 compared to the reference (30d window). This indicates that indeed it is important to capture death events occuring after end of progression follow-up. We can also see, that the effect of extending the window from 30d to 90d changes the HR by 0.11, suggesting that the effect of extending the window further is diminishing rapidly. This is reflected in the 'Incremental deaths' column as well: for the each 10-day period up to 30 days we capture approximately 300 more deaths, with this number dropping rapidly thereafter. It is hard to tell where exactly the cutoff should be made - but it looks like 30d to 60d is reasonable since most censoring-related deaths appear to be captured, and further extension has negligible impact on hazard ratios. 
  
<br/>  

# rwPFS in cancers other than aNSCLC, and custom filtering of progression events
  
<br/>

The above examples all used the aNSCLC datamart - how can rwPFS be calculated for other datamarts?

<br/>  
  
There are three points to consider:

- `fi_calc_rwPFS` and `fi_compare_rwPFS` can be used in exactly the same way as illustrated above. 

- Which rwPFS definitions are most appropriate must be determined individually for other patient populations. 

- `fi_add_progression` makes assumptions about the structure of the `PROGRESSION` table in the Flatiron database, which may complicate its use with certain datamarts. However, the function can be tweaked by using the `prog_filter_expression` argument. With it, a logical R expression (involving unquoted table column names) can be supplied, and only progression events are included where this expression evaluates to "TRUE". This can also be used for more fine-grained filtering of progression events in the aNSCLC datamart. 

<br/> 
 
For a start, let's look what the `PROGRESSION` table looks like in the aNSCLC datamart

<br/>

```{r, out.width = "90%", fig.align = "left"}
fk %>%
  fi_read_table(
    table_name = "ENHANCED_ADVNSCLCPROGRESSION"
  ) %>%
  head() 
```


<br/>

Knowing the structure of the progression table, we can now select radiographically confirmed progression events that are not pseudoprogression using the `prog_filter_expression` argument - which yields the same result as what we get by using the `require_radiographic` and `exclude_pseudoprogression` arguments.

<br/> 
 
```{r}
fk <- fk %>%
  #remove previously generated columns
  fi_select( 
    -contains("progressiondate"), 
    -contains("lastclinicnotedate"), 
    -contains("rwpfs")
  ) %>%
  
  #This is how we filtered progression events before
  fi_add_progression(
      start_date = "lot1startdate",          
      require_radiographic = TRUE,          
      exclude_pseudoprogression = TRUE,      
      discard_n_days = 0,                   
      label = "_radio_nopseudo",                  
      prog_filter_expression = NULL          
  ) %>%
  
  
  #This is the same using prog_filter_expression instead
  fi_add_progression(
      start_date = "lot1startdate",          
      require_radiographic = FALSE,          #These are now not used      
      exclude_pseudoprogression = FALSE,     #These are now not used
      discard_n_days = 0,                   
      label = "_prog_filter_expression",                  
      prog_filter_expression = "isradiographicevidence == 'Yes' &  ispseudoprogressionmentioned == 'No'"     
  )
  
  #This is to demonstrate that the two ways are exactly equivalent
  fk %>%
    fi_pull_data() %>%
    {
      all(.$progressiondate_radio_nopseudo == .$progressiondate_prog_filter_expression)
    }
  
```
  
<br/>  
As we can see, the two ways of filtering progression events are equivalent. 
<br/>

Note: require_radiographic or exclude_pseudoprogression arguments generate filter expressions that are simply concatenated with prog_filter_expression using logical `AND`. 

<br/>

To do this in Metastatic breastcancer, we run

<br/>
```{r}
fk2 <- fi_start("MetastaticBreastCancer")

fk2 %>%
  fi_read_table(table_name = "ENHANCED_METBREASTPROGRESSION") %>%
   head() 

```
  
<br/>

This progression table has additional columns/flags while containing the `ispseudoprogressionmentioned` and `isradiographicevidence` columns we've seen in the aNSCLC progression table. All these can be used jointly to filter progression events, for example
  
<br/>

```{r}
fk2 <- fk2 %>%
  fi_add_progression(
    start_date = "metdiagnosisdate",
    require_radiographic = FALSE,
    exclude_pseudoprogression = TRUE,
    discard_n_days = 0,
    label = "",
    prog_filter_expression = "istumormarkersonly == 'No'"
  )
```
  
<br/> 
  
In this manner we can use `fi_add_progression` with any progression table that has a `progressiondate` and a `lastclinicnotedate` column.

<br/>

```{r, include = FALSE}
fk$database$connection$disconnect()
fk2$database$connection$disconnect()
```
  
  

# References


<div id="refs"></div>
